Follow the Flow of Rain and Paint: \break An Output Sensitive Contour Tree Algorithm

### 2014-08-17 paint-contour-tree.tex
subsection5.2 added

@@

\pagebreak
\section{Lower Bound}
First we setup some basic terminology.
Any tree $T$ considered in this section will be a rooted full binary tree where the height of a vertex is its distance 
from the root $r$.  As such, the children of a vertex $v\in T$ are the adjacent vertices of larger height
(and $v$ is the parent of such vertices).  Then the subtree rooted at $v$, denoted $T_v$ consists of the graph induced on all 
vertices which are descendants of $v$ (including $v$ itself).  For two vertices $v$ and $w$ in $T$ let $d(v,w)$ denote the 
length of the path between $v$ and $w$.

@@

\begin{lemma}
\label{lem:algCost}
NEEDS TO BE WRITTEN:
Here we need a lemma showing that the running time of our algorithm is $O(f_{alg}(\chi(T)))$, where $T$ is the merge tree of the given input, 
and $\chi$ is any valid coloring of $T$.  (maybe add one in the definition of $H_v$ so that $\log |H_v|$ is always $>1$).
\end{lemma}

@@

\subsection{Path Decomposition Background}
We now introduce the concept of path decompositions and their cost.  In the next subsection we then relate this to the
running time of our algorithm.

@@

\begin{definition}
Let $P(T)$ be a path decomposition of $T$.  Then the paths in $P(T)$ naturally define a tree\footnote{Please excuse the 
overloading of the term 'tree', it is the most natural term to use here.} of there own.  Specifically, in the original 
tree $T$ contract each path down to its root.  Call the resulting tree the \emph{path tree} of $T$, corresponding to 
the path decomposition $P(T)$.

@@

THERE IS A PROBLEM in the following lemma.

@@

\subsection{Lower Bound by Path Decomposition}

\begin{lemma}
\label{lem:pathDecomp}
NEEDS TO BE WRITTEN:
Have a lemma here showing that our algorithm must pay at least the cost of any valid path decomposition.
\end{lemma}

For a given valid coloring $\chi$ of a tree $T$, we now describe a particular path decomposition for which 
we can prove $f_{alg}(\chi(T)) = O(f_{path}(P(T)))$ using the observations about decompositions from the 
previous subsection.  By the above lemma this will imply $f_{alg}(\chi(T)) = \Theta(f_{path}(P(T)))$.

>>>>

\newcommand{\XSays}[2]{{
      {$\rule[-0.12cm]{0.2in}{0.5cm}$\fbox{\tt
            #1:} }
      \textcolor{red}{#2}
      \marginpar{\textcolor{blue}{#1}}
      {$\rule[0.1cm]{0.3in}{0.1cm}$\fbox{\tt
            end}$\rule[0.1cm]{0.3in}{0.1cm}$}
      }
   }
\newcommand{\Ben}[1]{{\XSays{Ben}{#1}}}

@@

\pagebreak
\section{Lower Bound}

In this section we prove a lower bound on the time it takes to compute a merge tree.

\begin{definition}
 We say that a contour tree of a $2$-manifold is a \emph{merge tree} if it has a well defined root vertex such that height values 
 strictly increase along any root to leaf path.
\end{definition}

\begin{observation}
 The contour tree of an minimum dominant manifold (see \Def{dom}) is close to a merge tree.  Specifically, there is a root such that on any root 
 to leaf path height values strictly increase, with the only exception being potentially at the leaves themselves.  As such, up to constant 
 factors, any lower bound for computing the merge tree is a lower bound for computing the contour tree of a minimum dominant
 manifold (i.e. intuitively just clip off the leaves or shift them to lie above their parent).
\end{observation}

First we setup some basic terminology.
Any tree $T$ considered in this section will be a rooted full binary tree where the height of a vertex is its distance 
from the root $r$.  As such, the children of a vertex $v\in T$ are the adjacent vertices of larger height
(and $v$ is the parent of such vertices).  Then the subtree rooted at $v$, denoted $T_v$ consists of the graph induced on all 
vertices which are descendants of $v$ (including $v$ itself).  For two vertices $v$ and $w$ in $T$ let $d(v,w)$ denote the 
length of the path between $v$ and $w$.

@@

\Ben{The following lemma should be immediate once previous section is written properly.}
\begin{lemma}
\label{lem:algCost}
NEEDS TO BE WRITTEN:
Here we need a lemma showing that the running time of our algorithm is $O(f_{alg}(\chi(T)))$, where $T$ is the merge tree of the given input, 
and $\chi$ is any valid coloring of $T$.  (maybe add one in the definition of $H_v$ so that $\log |H_v|$ is always $>1$).
\end{lemma}

@@

\subsection{Path Decomposition Background}
\label{sec:background}
We now introduce the concept of path decompositions and their cost.  In the next subsection we then relate this to the
running time of our algorithm.

@@

\newcommand{\pathTree}{P_{\mathcal{T}}}
\begin{definition}
Let $P(T)$ be a path decomposition of $T$.  Then the paths in $P(T)$ naturally defines a tree\footnote{Please excuse the 
overloading of the term 'tree', it is the most natural term to use here.} of there own.  Specifically, in the original 
tree $T$ contract each path down to its root.  Call the resulting tree the \emph{path tree} of $T$ corresponding to 
the path decomposition $P(T)$, and denote it by $\pathTree(T)$.

@@

\Ben{There is a fundamental flaw in the following lemma and corollary.}

@@

\subsection{Lower Bound by Path Decomposition}
We say two path trees $T_1$ and $T_2$ (of merge trees of potentially different surfaces) are equivalent if there is an isomorphism, $f$,
between the two trees such that for any node $v\in T_1$ corresponding to a path of size $|p_v|$, $f(v)$ also corresponds to a path of size $|p_v|$.
For the purposes of the next lemma, we say two 2-manifolds are different, if their corresponding contour trees are the different.

\begin{figure}[h]\centering
    \includegraphics[width=.15\linewidth]{figs/diamond}\hspace{2cm}
    \includegraphics[width=.125\linewidth]{figs/topView}\hspace{2cm}
    \includegraphics[width=.095\linewidth]{figs/sideView}
    \caption{From left to right: the cone portion of a diamond; the top down view and then the side view of a cone face with a platform added.}
    \label{fig:diamond}
\end{figure}

\begin{lemma}
\label{lem:pathDecomp}
Let $T$ be a merge tree, $P = P(T)$ any path decomposition of $T$, and $\pathTree$ the corresponding path tree.  
Then there are at least $\Pi_{p_i\in P} |p_i|!$ different 2-manifolds with an equivalent path tree.
\end{lemma}
\begin{proof}
 We describe a simple input mesh, with heights specified for all vertices except for the saddles.  The main idea is that 
 this mesh will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of merge trees 
 whose size is that described in the lemma statement.

 First, we describe the basic building block of this input mesh, which roughly speaking corresponds to a fixed path $p\in P$.  
 Specifically, we say a \emph{diamond} of size $m$ is the following mesh.  To start, construct a cycle 
 of length $m$, where the height of the $i$th vertex in the cycle is say $i/2m$.  Next, create an apex vertex at height 
 $\infty$ (or any appropriately large finite height), and add an edge between this vertex and each vertex in the cycle.  At this point 
 we have built an upside down cone with $m$ triangular faces (see \Fig{diamond}). To make this cone a closed surface we add a triangular cap at the bottom. 
 Specifically, add a $3$ cycle of vertices at heights $0$, $1/20m$, and $1/10m$, and connect each vertex of this $3$-cycle to an arbitrary distinct 
 vertex on the $m$ cycle (with matching cyclic ordering), and triangulate all remaining faces.  For consistency, if the size $m$ is $\leq 3$, the diamond 
 will be just a tetrahedron.

 Now above each face of the cone of the diamond, we will create a triangular ``platform'', which will act as the location to glue on other (shrunken) diamonds, 
 and whose highest point of connection to the face below will be a saddle vertex.  Specifically the heights of the vertices of the triangle will be $h$, $h-\eps$, and $h-2\eps$, where 
 $h$ is some unspecified height value.  This near horizontal platform is then connected to its respective face with vertical edges down to the face. See \Fig{diamond}.
 %
 %The vertex at height $h$ will lie in the plane of the respective face of the cone, and the other two will lie above the face, 
 %with edges connecting down to the face, hence creating an actual ``platform''.

 At this point we can construct the entire input mesh, by gluing together diamonds.  Specifically, for each path $p_i\in P$ we create a diamond of size $p_i$.
 Now we glue together diamonds of different paths in the same way the paths are connected in $\pathTree$.  Specially, for two paths $p,q\in P$ where 
 $p$ is the parent of $q$ in $\pathTree$, we glue the bottom triangular cap of the diamond for $q$ onto the triangular ``platform'' of some face of the diamond for $p$.
 (Naturally for this construction to work, diamonds for a given path will be scaled down relative to the size of the diamond of their parent).

 Observe now that the only saddle points in this construction are the highest connecting point on a face of an edge from the platform above.  Moreover, the only maxima are the apexes of the 
 diamonds, and there is exactly one minima corresponding to the bottom cap of the diamond representing the root of $\pathTree$.  Therefore, the saddles on a given 
 diamond will appear contiguously on a root to leaf path in the merge tree of the mesh, where the leaf corresponds to the maxima of the diamond 
 (since all these saddles have a direct line of sight to this apex).  In particular, 
 this implies that, regardless of the heights assigned to the platforms, the merge tree of this mesh has a path decomposition whose corresponding path 
 tree is equivalent to $\pathTree$.

 Now there is a valid instantiation of this described mesh for any relative ordering of the heights of the saddles on a given diamond.  In particular, 
 there are $|p|!$ possible orderings of the heights of the saddles on the diamond for $p$, and hence $\Pi_{p_i\in P} |p_i|!$ possible instantiations of 
 the mesh overall.  Also, each one of these instantiations will result in a different merge.  Specifically, a permutation of the heights on a given diamond 
 corresponds to a permutation of the vertices along a path in $P$, as by construction all saddles on a given diamond will appear in sorted order in the merge tree.
\end{proof}

\begin{corollary}
 Let $A(S)$ be any algorithm to compute the merge tree of any input surface $S$ (whose contour tree is a merge tree).  
 Then the running time of $A(S)$ is $\Omega(f_{path}(P))=\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$ where $P$ is any path decomposition of the 
 merge tree of $S$.
\end{corollary}
\begin{proof}
 Given an input surface $S$ whose merge tree has a particular path decomposition $P$, by the above lemma, there are $\Pi_{p_i\in P} |p_i|!$ other surfaces 
 with distinct merge trees, all of which have a path tree equivalent to $\pathTree$.  Therefore, any algorithm which computes the merge tree must distinguish between 
 these possibilities, and therefore its corresponding decision tree must have $\Pi_{p_i\in P} |p_i|!$ leaves.  Hence by Stirling's approximation, 
 this decision tree must have depth $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
\end{proof}

\subsection{Upper Bound by Path Decomposition}
\Ben{This section needs revision.}

From the previous section we now have a lower bound of $\Omega(f_{path}(P))$ for computing the merge tree in terms of any path decomposition $P$. 
We now show that the running time of our algorithm for computing the merge tree is upper bounded by this expression for a particular path decomposition.

Specifically, for a given valid coloring $\chi$ of a tree $T$, we describe a path decomposition for which 
we can prove $f_{alg}(\chi(T)) = O(f_{path}(P(T)))$ using the observations about path decompositions from 
\Sec{background}, which will then imply $f_{alg}(\chi(T)) = \Theta(f_{path}(P(T)))$.
